---
title: "Bellabeat"
author: "Paul Hellinger"
date: "2025-01-22"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Prepare the Data:

Download and install packages

```{r}
install.packages('tidyverse')
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
```

Set working directory and upload files

```{r}
setwd("~/Desktop/FitBitData")
activity <- read.csv("dailyActivity_merged4.12-5.12.csv")
sleep <- read.csv("sleepDay_merged.csv")
weight <- read.csv("weightLogInfo_merged.csv")
hourly_steps <- read.csv("hourlySteps_merged.csv")
hourly_calories <- read.csv("hourlyCalories_merged.csv")
hourly_intensities <- read.csv("hourlyIntensities_merged.csv")
```

Review the Data: Take a brief look at the datasets

```{r}
head(activity)
head(sleep)
head(weight)
head(hourly_steps)
head(hourly_calories)
head(hourly_intensities)
str(activity)
str(sleep)
str(weight)
str(hourly_steps)
str(hourly_calories)
str(hourly_intensities)
```

We see that the top three data sets all have varying column names for
their dates and they are in different formats.

Clean the Data: Check for any duplicate entries.

```{r}
# Check for duplicates in activity 
activity_duplicates <- activity %>% filter(duplicated(.))
print(activity_duplicates)

# Check for duplicates in sleep 
sleep_duplicates <- sleep %>% filter(duplicated(.))
print(sleep_duplicates)

# Check for duplicates in weight 
weight_duplicates <- weight %>% filter(duplicated(.))
print(weight_duplicates)
```

Remove duplicate entries in sleep

```{r}
# Remove duplicates from sleep
sleep_clean <- sleep %>% distinct()
```

Verify data has been cleaned properly:

```{r}
# Verify that duplicates are removed
any(duplicated(sleep_clean))    # Should return FALSE
```

Format the various date columns in activity, sleep_clean and weight to
proper formats:

```{r}
# Add Date column to activity and delete ActivityDate
activity <- activity %>%
  mutate(Date = as.Date(ActivityDate, format = "%m/%d/%Y")) %>%
  select(-ActivityDate)

# Add Date column to sleep_clean and delete SleepDay
sleep_clean <- sleep_clean %>%
   mutate(Date = as.Date(SleepDay, format = "%m/%d/%Y")) %>%
  select(-SleepDay)

# Change Date column to proper format
weight <- weight %>%
   mutate(Date = as.Date(Date, format = "%m/%d/%Y")) 
 
```

Format date to proper format and remove excess columns for hourly data:

```{r}

# Add Date column and hour column and remove ActivityHour from hourly_steps
hourly_steps <- hourly_steps %>%
  mutate(
    ActivityHour = mdy_hms(ActivityHour),  # Convert to datetime
    Hour = hour(ActivityHour)             # Extract the hour (0–23 format)
  ) %>%
  mutate(
    Date = as.Date(ActivityHour, format = "%m/%d/%Y")
  )
# Remove excess column
hourly_steps <- hourly_steps %>% select(-ActivityHour)

# Add Date column and hour column and remove ActivityHour from hourly_calories
hourly_calories <- hourly_calories %>%
  mutate(
    ActivityHour = mdy_hms(ActivityHour),  # Convert to datetime
    Hour = hour(ActivityHour)             # Extract the hour (0–23 format)
  ) %>%
  mutate(
    Date = as.Date(ActivityHour, format = "%m/%d/%Y")
  )
# Remove excess column
hourly_calories <- hourly_calories %>% select(-ActivityHour)


# Add Date column and hour column and remove ActivityHour from hourly_intensities
hourly_intensities <- hourly_intensities %>%
  mutate(
    ActivityHour = mdy_hms(ActivityHour),  # Convert to datetime
    Hour = hour(ActivityHour)             # Extract the hour (0–23 format)
  ) %>%
  mutate(
    Date = as.Date(ActivityHour, format = "%m/%d/%Y")
  )
# Remove excess column
hourly_intensities <- hourly_intensities %>% select(-ActivityHour)

```

Combine datasets:

```{r}
# Merge two daily data sets and then merge again with the third set
activitySleep <- activity %>%
  full_join(sleep_clean, by = c("Id","Date"))
daily_data <- activitySleep %>%
  full_join(weight, by = c("Id", "Date"))

# merge hourly data sets
caloriesSteps <- hourly_calories %>%
  full_join(hourly_steps, by = c("Id","Date","Hour"))
hourly_data <- caloriesSteps %>%
  full_join(hourly_intensities, by = c("Id","Date","Hour"))

```

Adding a column for the day of the week to both sets will be helpful
when analyzing the data later:

```{r}

# Add Weekday column to daily_data and format them in proper order for plotting later
daily_data <- daily_data %>%
  mutate(
    Weekday = weekdays(Date),  # Add Weekday column
    Weekday = factor(Weekday, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))  # Specify order
  )

# Add Weekday column to hourly_data and format them in proper order for plotting later
hourly_data <- hourly_data %>%
  mutate(
    Weekday = weekdays(Date),  # Add Weekday column
    Weekday = factor(Weekday, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))  # Specify order
  )

# Add TotalMinutes column. This will be important later for plotting.
daily_data <- daily_data %>%
  mutate(TotalMinutes = VeryActiveMinutes + FairlyActiveMinutes + LightlyActiveMinutes + SedentaryMinutes)


```

Save the Data:

```{r}
# Save the two datasets for upload to Tableau later
write_csv(daily_data, "daily_data.csv") # Save daily_data dataset
write_csv(hourly_data, "hourly_data.csv") # Save hourly_data dataset
```

Examine the Data: We are told there are only 30 participants in this
survey. Lets take a look at the number of unique Ids in the hourly_data
and daily_data

```{r}
# Check for the number of unique Ids in hourly_data and daily_data
n_distinct(hourly_data$Id)
n_distinct(daily_data$Id)
```

For both we get a return of 33, 3 more than we were expecting. We
noticed before while we were briefly looking over the data that the
datasets sleep_clean and weight contain much few inputs than activity.
Lets find the unique users in all three sets: activity, sleep_clean and
weight

```{r}
# Check for the number of unique Ids in activity, sleep_clean and weight
n_distinct(activity$Id)
n_distinct(sleep_clean$Id)
n_distinct(weight$Id)
```

We get a return of 33, 24 and 8. This means activity has 3 additional
unexpected Ids, sleep_clean has 6 fewer Ids and weight has 22 fewer Ids
than expected. The sleep_clean and weight datasets contain Ids fewer
than minimum sample sizes, but they can still provide useful insights
into how the participants were using their trackers and smart devices.

Let's first examine the overall activity of the users with the largest
scope, meaning let's look at the daily_data. We will break down the data
by weekday.

```{r}

# Examine the average steps by weekday
daily_trends <- daily_data %>%
  group_by(Weekday) %>%
  summarise(
    AvgSteps = mean(TotalSteps, na.rm = TRUE),
    AvgCalories = mean(Calories, na.rm = TRUE),
    AvgDistance = mean(TotalDistance, na.rm = TRUE)
  )
# Plot average steps by weekday
ggplot(daily_trends, aes(x = Weekday, y = AvgSteps, fill = Weekday)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Steps by Weekday", x = "Weekday", y = "Average Steps") +
  theme_minimal()

# Plot average calories burned by weekday
ggplot(daily_trends, aes(x = Weekday, y = AvgCalories, fill = Weekday)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Calories Burned by Weekday", x = "Weekday", y = "Average Calories") +
  theme_minimal()

# Plot average distance traveled by weekday
ggplot(daily_trends, aes(x = Weekday, y = AvgDistance, fill = Weekday)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Distance Traveled by Weekday", x = "Weekday", y = "Average Distance") +
  theme_minimal()

```

We see that the plot for distance traveled and steps correspond 1:1 to
each other, which is expected. Interestingly, we see that there isn't
necessarily a relationship between calories burned and steps taken,
specifically when it comes to Monday and Friday. These days might
involve exercieses that burn calories but don't correspond with a high
step count, weight lifting or body resistance exercises for example.

Instead of looking at averages, let's look at totals.

```{r}
# Aggregate the data by Weekday and calculate the total for steps, distance, and calories
daily_totals <- daily_data %>%
  group_by(Weekday) %>%
  summarise(
    TotalSteps = sum(TotalSteps, na.rm = TRUE),
    TotalCalories = sum(Calories, na.rm = TRUE)
  )

# Plot the totals for each day of the week (you can create one plot per metric or combine them)
ggplot(daily_totals, aes(x = Weekday)) +
  geom_bar(aes(y = TotalSteps, fill = "Steps"), stat = "identity", position = "dodge")  +
  geom_bar(aes(y = TotalCalories, fill = "Calories"), stat = "identity", position = "dodge") +
  labs(
    title = "Total Steps and Calories Burned by Weekday",
    x = "Weekday",
    y = "Total"
  ) +
  scale_fill_manual(values = c("Steps" = "blue", "Calories" = "red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This plot shows an interesting over-looked point. The total calories
burned and steps tracked are largest between tuesday and thursday. We
should examine the data further to see how often users tracked their
data and on what days.

```{r}
# Count the number of days each user inputted data
user_data_summary <- daily_data %>%
  group_by(Id) %>%
  summarise(
    DaysTracked = n_distinct(Date),  # Count unique dates per user
    FirstDate = min(Date),          # First date a user inputted data
    LastDate = max(Date)            # Last date a user inputted data
  ) 

# Count inputs by weekday
weekday_summary <- daily_data %>%
  group_by(Weekday) %>%
  summarise(
    Entries = n(),                       # Total number of entries
  ) %>%
  arrange(match(Weekday, c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))

# Plot data
ggplot(weekday_summary, aes(x = Weekday, y = Entries, fill = Weekday)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Inputs by Weekday",
    x = "Weekday",
    y = "Number of Inputs",
    fill = "Weekday"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

The generated plot shows what was expected: there are more inputs during
the middle of the week from Tuesday to Thursday than any other days of
the week, thus users are more likely to input data during Tuesday
through Thursday than other days. Finding a way to promote usage of
smart devices during the other days of the week will lead to healthier
lifestlye choices.

We will plot the total calories v the total steps. Each blue dot will represent the totals for a specific Id on a specific date. 
```{r}
# Create the scatterplot with trend line
ggplot(daily_data, aes(x = TotalSteps, y = Calories)) +
  geom_point(color = "blue") +  # Scatter plot points in blue
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear trend line in red, without confidence interval
  labs(
    title = "Total Calories vs Total Steps",
    x = "Total Steps",
    y = "Total Calories"
  ) +
  theme_minimal()  # Minimal theme for cleaner look
```
There's an expected result: The higher the total steps, the higher the total calories burned.

Now let's change gears and take a look at the sleep_clean data set and
examine it further:

```{r}

# Add Weekday column
sleep_clean <- sleep_clean %>%
  mutate(
    Weekday = weekdays(Date),
    Weekday = factor(Weekday, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))  # Specify order
    )

# Convert to hours 
sleep_clean <- sleep_clean %>%
  mutate(
    TotalHoursAsleep = TotalMinutesAsleep / 60,
    TotalHoursInBed = TotalTimeInBed / 60
  )

# Examine how much sleep the participants are having on average
user_sleep_summary <- sleep_clean %>%
  group_by(Id) %>%
  summarize(
    AvgHoursAsleep = mean(TotalHoursAsleep, na.rm = TRUE),
    AvgHoursInBed = mean(TotalHoursInBed, na.rm = TRUE),
    .groups = "drop"
  )
print(user_sleep_summary)

# Examine sleeping habits on different days of the week 
sleep_weekday_summary <- sleep_clean %>%
  group_by(Weekday) %>%
  summarize(
    AvgHoursAsleep = mean(TotalHoursAsleep, na.rm = TRUE),
    AvgHoursInBed = mean(TotalHoursInBed, na.rm = TRUE),
    .groups = "drop"
  )

print(sleep_weekday_summary)

# Calculate how many days each user tracks their sleep
user_sleep <- sleep_clean %>%
  group_by(Id) %>%
  summarize(DaysTracked = n(), .groups = "drop")

print(user_sleep)

# Calculate average DaysTracked
avg_days_tracked <- user_sleep %>%
  summarize(AverageDaysTracked = mean(DaysTracked, na.rm = TRUE))

print(avg_days_tracked)
```

Now that we've created these tables, we'll plot them for further
insights

```{r}

# Create a plot of the average hours slept by weekday
ggplot(sleep_weekday_summary, aes(x = Weekday, y = AvgHoursAsleep)) +
  geom_col(fill = "lightgreen") +
  labs(
    title = "Average Hours Slept by Weekday",
    x = "Day of the Week",
    y = "Average Hours Asleep"
  ) +
  theme_minimal()

```

Let's examine further by combining the data on the number of days
tracked and averages of the amount of sleep and time in bed for all participants. 

```{r}
# Combine data sets
sleep_combined <- user_sleep %>%
  full_join(user_sleep_summary, by = c("Id"))
```

Now examine the data to see if there is a relationship between number of
days tracked and the amount of sleep the users got.

```{r}
sleep_plot <- ggplot(sleep_combined, aes(x = reorder(Id, -DaysTracked), y = DaysTracked, fill = AvgHoursAsleep)) + ## Reorder Ids from most days tracked to least
  geom_col() +
  scale_fill_gradient(low = "lightgreen", high = "darkblue", name = "Avg Hours Asleep") +
  labs(
    title = "Days Tracked and Average Hours Asleep by User",
    x = "User ID",
    y = "Days Tracked"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sleep_plot

```

As a whole, we can see that there is a positive relationship between the
number of days tracked and average hours slept. We also see that on
average users only inputted 17 days out of the 31 total. We can
recommend a marketing stratedgy focused on tracking sleep to promote
healthier sleeping habits.

Now let's take a look at the hourly data.

```{r}

# Find averages per Id and Hour
hourly_summary <- hourly_data %>%
  group_by(Id, Hour) %>%
  summarize(
    AverageSteps = mean(StepTotal, na.rm = TRUE),
    AverageCalories = mean(Calories, na.rm = TRUE),
    AverageIntensity = mean(AverageIntensity, na.rm = TRUE),
    TotalEntries = n(),  # Count total entries for consistency tracking
    .groups = "drop"
  )

# Aggregate hourly_summary data across all users
hourly_overall_summary <- hourly_summary %>%
  group_by(Hour) %>%
  summarize(
    AvgStepsAllUsers = mean(AverageSteps, na.rm = TRUE),
    AvgCaloriesAllUsers = mean(AverageCalories, na.rm = TRUE),
    AvgIntensityAllUsers = mean(AverageIntensity, na.rm = TRUE),
    .groups = "drop"
  )

# Normalize values for easier comparison
hourly_overall_summary <- hourly_overall_summary %>%
  mutate(
    ScaledSteps = AvgStepsAllUsers / max(AvgStepsAllUsers, na.rm = TRUE),
    ScaledCalories = AvgCaloriesAllUsers / max(AvgCaloriesAllUsers, na.rm = TRUE),
    ScaledIntensity = AvgIntensityAllUsers / max(AvgIntensityAllUsers, na.rm = TRUE)
  )

# Plot scaled values for comparison
library(ggplot2)

ggplot(hourly_overall_summary, aes(x = Hour)) +
  geom_line(aes(y = ScaledSteps, color = "Steps")) +
  geom_line(aes(y = ScaledCalories, color = "Calories")) +
  geom_line(aes(y = ScaledIntensity, color = "Intensity")) +
  labs(
    title = "Comparison of Steps, Calories, and Intensity Across All Users (Scaled)",
    x = "Hour of the Day",
    y = "Scaled Average Value",
    color = "Metric"
  ) +
  theme_minimal()

```

From the created plot, we see a direct correlation between all three
averages: steps, intensities, and calories burned. The more steps, the
higher the intensity and the more calories burned.


Now let's aggregate the data so that it factors in days of the week and
create a heatmap from the data.

```{r}

# Summarize metrics by Weekday and Hour
hourly_weekday_summary <- hourly_data %>%
  group_by(Weekday, Hour) %>%
  summarize(
    AvgSteps = mean(StepTotal, na.rm = TRUE),
    AvgCalories = mean(Calories, na.rm = TRUE),
    AvgIntensity = mean(TotalIntensity, na.rm = TRUE),
    .groups = "drop"
  )

# Heatmap for steps with angled x-axis labels
ggplot(hourly_weekday_summary, aes(x = Hour, y = Weekday, fill = AvgSteps)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Avg Steps") +
  labs(
    title = "Average Steps by Hour and Weekday",
    x = "Hour of the Day",
    y = "Weekday"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Angled x-axis labels
  )

```

We see from the heatmap some habits that we would expect: users appear
to be sleeping in on saturday/sunday and we see a general change in the
block from 6 am to 8 pm on the weekdays compared to Saturday and Sunday.
Saturday also appears to generally be more active during the day while
also being a day the users had more sleep. An unexpected habit is the
high average step count on Wednesday afternoon between 5 and 7. This
appears to be the most active time on average for all users. Also,
saturday afternoon appears to be the second most active time. To promote
healthier lifestyle habits, Bellabeat marketing team could focus on
increasing the average step count on Monday and Friday, creating a week
with more consistent active days.

Now let's examine a bit further, looking at the minute data that we
have. It is embedded in the daily_data dataset. We added a total minutes
column earlier so we can see how the different activity levels compare
to each other.

```{r}

# Add percentage calculation for the legend
activity_summary <- daily_data %>%
  summarise(
    VeryActive = sum(VeryActiveMinutes, na.rm = TRUE),
    FairlyActive = sum(FairlyActiveMinutes, na.rm = TRUE),
    LightlyActive = sum(LightlyActiveMinutes, na.rm = TRUE),
    Sedentary = sum(SedentaryMinutes, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "ActivityLevel", values_to = "TotalMinutes") %>%
  mutate(
    Percentage = TotalMinutes / sum(TotalMinutes) * 100,
    Label = paste0(ActivityLevel, " (", round(Percentage, 1), "%)") # Combine label/percentage
  )
```

Now plot the data.

```{r}
# Create the pie chart
ggplot(activity_summary, aes(x = "", y = TotalMinutes, fill = Label)) +
  geom_bar(stat = "identity", width = 1, color = "white") + # Add white borders
  coord_polar("y", start = 0) + # Convert to pie chart
  labs(
    title = "Distribution of Activity Levels by Total Minutes",
    fill = "Activity Level"
  ) +
  theme_void() +
  theme(
    legend.position = "right", # Move legend to the right
    legend.title = element_text(face = "bold", size = 12), # Style the legend title
    legend.text = element_text(size = 10), # Style legend text
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold") # Center the title
  )
```

We see out of the total time tracked, around 3 percent of the time can
be considered very or fairly active. According to the CDC, adults should
get about 30 minutes a day of moderately intense physical activity. This
is about 2 percent of the total day. That means that on total, the
participants are getting around the recommended amount of exercise per
day. It's interesting to note that VeryActive was slightly larger than
FairlyActive, which is not expected.

